Exoplanet population inference and the abundance \\ of Earth analogs from noisy, incomplete catalogs

### 2014-08-23 document/ms.tex
summarizing

In the next Section, we introduce our method for incorporating the effects of
observational uncertainties into the derivation of occurrence rates.
The technical term for this procedure is \emph{hierarchical inference} and
while a general discussion of this field is beyond the scope of this \paper,
in the next section, we present the basic probabilistic question and derive a
computationally tractable method for inferring the population of exoplanets.
In \sect{model}, we explain how to incorporate the effects of survey
completeness into our inferences and derive the key equation for our method:
\eq{money}.
We test our method on synthetic catalogs in Sections~\sectalt{valid}
and~\sectalt{extrap}.
In \sect{real}, we use the catalog of planet candidates and the empirically
determined detection efficiency from \citet{petigura} to measure the
occurrence rate density of small planets with long orbital periods.

@@

\section{A brief introduction to hierarchical inference}

In this \paper, we are asking the question: \emph{what constraints can we put
on the occurrence rate density of exoplanets given all the light curves
measured by \kepler?}
Throughout this \paper, we use the notation $\rate_\ratepars(\entry)$ for the
occurrence rate density \rate---parameterized by the parameters \ratepars---as
a function of the physical parameters \entry\ (orbital period, planetary
radius, \etc).
In this framework, the occurrence rate density can be ``parametric''---for
example, a power law---or a ``non-parametric'' function---such as a histogram
where the bin heights are the parameters \ratepars.
The light curve for a particular target $k$ is $\data_k$ and the \True\ set of
physical parameters for that same object is given by $\entry_k$.
A catalog provides a set of constraints on the parameters $\{\entry_k\}$
conditioned on the set light curves $\{\data_k\}$.

Any inference about the rate density parameters can be formally addressed by
evaluating the \emph{marginalized likelihood function}
\begin{eqnarray}\eqlabel{crazylike}
p(\{\data_k\}\,|\,\ratepars) &=&
    \int p(\{\data_k\}\,|\,\{\entry_k\})
    \,p(\{\entry_k\}\,|\,\ratepars)
    \dd\{\entry_k\}
\end{eqnarray}
where, as we mentioned above, $\{\data_k\}$ is the set of all light curves,
one light curve $\data_k$ per target $k$, \ratepars\ is the vector of
parameters describing the population occurrence rate density
$\rate_\ratepars(\entry)$ and $\entry_k$ is the vector of physical parameters
describing the planetary system (orbital periods, radius ratios, stellar
radius, \etc) around target $k$.
In this equation, our only assumption is that the datasets depend on the
rate density of exoplanets only through the catalog $\{\entry_k\}$.
In our case, this assumption qualitatively means that the signals found in the
light curves depend only on the actual properties of the planet and star, and
not on the distributions from which they are drawn.
It is worth emphasizing that---as we will discuss further below---the catalog
only provides \emph{probabilistic constraints} on $\{\entry_k\}$; not perfect
delta-function measurements.

In other words, we treat a catalog as being a dimensionality reduction of the
raw data with all the relevant information retained.
In the context of \kepler, the catalog reduces the set of downloaded pixel
time series (approximately 70,000 data points for the typical \kepler\ target)
to probabilistic constraints on a handful of physical parameters---\entry\
from above---like the orbital period and planetary radius.
If we take this set of parameters $\{\entry_k\}$ as \emph{sufficient
statistics} of the data then we can, in theory, compute \eq{crazylike}---up to
an unimportant constant---without ever looking at the raw data again!
This is important because the high-dimensional integral in \eq{crazylike}
won't generally have an analytic solution and each evaluation of the
per-object likelihood $p(\data_k\,|\,\entry_k)$ is expensive, making numerical
methods intractable.

@@

Now, we can substitute \eq{poisson-like} into \eq{simplemarglike} and apply
the importance sampling approximation (\eqalt{importance}) to derive the
following expression for the marginalized likelihood
\begin{eqnarray}\eqlabel{money}
\frac{p(\{\data_k\}\,|\,\ratepars)}{p(\{\data_k\}\,|\,\interim)} &\approx&
    \exp\left(-\int \obs{\rate}_\ratepars (\entry) \dd\entry\right) \,
    \prod_{k=1}^K
    \frac{1}{N_k} \sum_{n=1}^{N_k}
    \frac{\obs{\rate}_\ratepars (\entry_k^{(n)})}
         {p(\entry_k^{(n)}\,|\,\interim)} \quad.
\end{eqnarray}
In this equation, we're making the further assumption that the catalog treated
the objects independently.
This is a somewhat subtle point if we were to consider targets with more than
one transiting planet---a point that we will return to below---but for the
considerations of the dataset considered in this \paper, it is a justified
simplification.
In other words, we have per-object posterior samples
\begin{eqnarray}
\entry_k^{(n)} &\sim& p(\entry_k\,|\,\data_k,\,\interim) \quad.
\end{eqnarray}
\Eq{money} is the \emph{money equation} for our method.
It lets us efficiently compute the \emph{marginalized likelihood of the entire
set of light curves for a particular occurrence rate density}.

\paragraph{Occurrence rate density parameterization \& priors}
For the remainder of this \paper, we also model the rate density as a
two-dimensional histogram with fixed logarithmic bins in period and radius.
When we include observational uncertainties---using \eq{money}---the maximum
likelihood result is no longer analytic; \eq{ml-ide} is no longer applicable
when the uncertainties are non-negligible.
Therefore, if we want to compute the ``best-fit'' rate density, we can use a
standard non-linear optimization algorithm.
In the regions of parameter space that we tend to care about, the completeness
is low and there are only a few observations with large uncertainties.
In this case, we're especially interested in probabilistic constraints on the
occurrence rate density; not just the best-fit model.
To do this, we must apply a prior $p(\ratepars)$ on the rate density
parameters and generate samples from the posterior probability
\begin{eqnarray}\eqlabel{posterior}
p(\ratepars\,|\,\{\data_k\}) &\propto&
    p(\ratepars)\,p(\{\data_k\}\,|\,\ratepars)
\end{eqnarray}
using Markov chain Monte Carlo (MCMC).

>>>>

In the next Section, we summarize the likelihood method for exoplanet
population inference and in \sect{hier}, we describe how to include the
effects of observational uncertainties.
The technical term for this procedure is \emph{hierarchical inference} and
while a general discussion of this field is beyond the scope of this \paper,
in the \sect{hier}, we present the basic probabilistic question and derive a
computationally tractable inference procedure.
In \sect{model}, we summarize the technique and derive the key equation for
our method: \eq{money}.
We test our method on synthetic catalogs in Sections~\sectalt{valid}
and~\sectalt{extrap}.
In \sect{real}, we use the catalog of planet candidates and the empirically
determined detection efficiency from \citet{petigura} to measure the
occurrence rate density of small planets with long orbital periods.

@@

\section{A brief introduction to hierarchical inference}
\sectlabel{hier}

The general question that we are trying to answer in this \paper\ is:
\emph{what constraints can we put on the occurrence rate density of exoplanets
given all the light curves measured by \kepler?}
In the case of negligible measurement uncertainties, this is equivalent to
optimizing \eq{poisson-like} but when this approximation is no longer valid,
we must instead compute the \emph{marginalized likelihood}
\begin{eqnarray}\eqlabel{crazylike}
p(\{\data_k\}\,|\,\ratepars) &=&
    \int p(\{\data_k\}\,|\,\{\entry_k\})
    \,p(\{\entry_k\}\,|\,\ratepars)
    \dd\{\entry_k\}
\end{eqnarray}
where $\{\data_k\}$ is the set of all light curves, one light curve $\data_k$
per target $k$, \ratepars\ is the vector of parameters describing the
population occurrence rate density $\rate_\ratepars(\entry)$ and $\entry_k$ is
the vector of physical parameters describing the planetary system (orbital
periods, radius ratios, stellar radius, \etc) around target $k$.
In this equation, our only assumption is that the datasets depend on the
rate density of exoplanets only through the catalog $\{\entry_k\}$.
In our case, this assumption qualitatively means that the signals found in the
light curves depend only on the actual properties of the planet and star, and
not on the distributions from which they are drawn.
It is worth emphasizing that---as we will discuss further below---the catalog
only provides \emph{probabilistic constraints} on $\{\entry_k\}$; not perfect
delta-function measurements.

In other words, we treat the catalog as being a dimensionality reduction of
the raw data with all the relevant information retained.
In the context of \kepler, the catalog reduces the set of downloaded pixel
time series (approximately 70,000 data points for the typical \kepler\ target)
to probabilistic constraints on a handful of physical parameters---\entry\
from above---like the orbital period and planetary radius.
If we take this set of parameters $\{\entry_k\}$ as \emph{sufficient
statistics} of the data then we can, in theory, compute \eq{crazylike}---up to
an unimportant constant---without ever looking at the raw data again!
This is important because the high-dimensional integral in \eq{crazylike}
won't generally have an analytic solution and each evaluation of the
per-object likelihood $p(\data_k\,|\,\entry_k)$ is expensive, making numerical
methods intractable.

@@

Now, we can substitute \eq{poisson-like} into \eq{simplemarglike} and apply
the importance sampling approximation (\eqalt{importance}) to derive the
following expression for the marginalized likelihood
\begin{eqnarray}\eqlabel{money}
\frac{p(\{\data_k\}\,|\,\ratepars)}{p(\{\data_k\}\,|\,\interim)} &\approx&
    \exp\left(-\int \obs{\rate}_\ratepars (\entry) \dd\entry\right) \,
    \prod_{k=1}^K
    \frac{1}{N_k} \sum_{n=1}^{N_k}
    \frac{\obs{\rate}_\ratepars (\entry_k^{(n)})}
         {p(\entry_k^{(n)}\,|\,\interim)}
\end{eqnarray}
where the values $\{{\entry_k}^{(n)}\}$ are samples drawn from the posterior
probability
\begin{eqnarray}
{\entry_k}^{(n)} &\sim& p(\entry_k\,|\,\data_k,\,\interim)
\end{eqnarray}
as described in the previous section.
\Eq{money} is the \emph{money equation} for our method.
It lets us efficiently compute the \emph{marginalized likelihood of the entire
set of light curves for a particular occurrence rate density}.

In this equation, we're making the further assumption that the catalog treated
the objects independently.
This is a somewhat subtle point if we were to consider targets with more than
one transiting planet---a point that we will return to below---but for the
considerations of the dataset considered here, it is a justified
simplification.

For the remainder of this \paper, we model the rate density as a
two-dimensional histogram with fixed logarithmic bins in period and radius.
When we include observational uncertainties---using \eq{money}---the maximum
likelihood result is no longer analytic.
Therefore, if we want to compute the ``best-fit'' rate density, we can use a
standard non-linear optimization algorithm.

In the regions of parameter space that we tend to care about, the completeness
is low and there are only a few observations with large uncertainties.
In this case, we're especially interested in probabilistic constraints on the
occurrence rate density; not just the best-fit model.
To do this, we must apply a prior $p(\ratepars)$ on the rate density
parameters and generate samples from the posterior probability
\begin{eqnarray}\eqlabel{posterior}
p(\ratepars\,|\,\{\data_k\}) &\propto&
    p(\ratepars)\,p(\{\data_k\}\,|\,\ratepars)
\end{eqnarray}
using Markov chain Monte Carlo (MCMC).
